Exploring the Soil Microbiome coupling with the Crete island Data Cube

### 2024-04-06 manuscript.tex
update abstract to 250 words

\abstract{Ecosystem functioning is an integral part of the sciences of climate
    change and conservation. Microbes are known for their versatility, abundance
    and influence on ecosystem functions, but a more thorough analysis of
    ecosystems should include a more complex assembly of organisms.
    For example, plant-arthropod-soil microbiome interactions are important
    associations that have historically been ignored. A synthesized knowledge
    base of biodiversity, in terms of ecological and remote-sensing data remains a
    major challenge. Many worldwide studies have been published regarding soil
    microbiome ecosystems, though there are many gaps to cover the complexity
    of functioning and biodiversity in the local scale. Islands can be
    important case studies for this integration, from micro to macro, with Crete being a great example. 
    Here, we utilize the Island Sampling Day Crete 2016 microbial 16S rRNA gene
    amplicon data and metadata, integrated with soil, faunistic and remote
    sensing data, to decipher the drivers of ecosystem function of the island.
    Cretan macroecology has been studied for centuries for its diverse and
    unique geology, fauna and flora. In addition, Crete as a continental island,
    presents a distinct natural and evolutionary history with high contrasts in
    vegetation cover, climatic conditions and geology. The Island Sampling Day
    Crete 2016 project (co-hosted by the Genomic Standards Consortium (GSC) and
    the Institute of Marine Biology, Biotechnology and Aquaculture (IMBBC-HCMR)), has
    collected microbial 16s amplicon data from 72 distinct, with ecosystem
    diversity topsoil sites from all around Crete, accompanied by FAIR
    (Findable, Accessible, Interoperable and Reproducible) data by design. 
    With this island wide study the GSC put the genomic standards in action.
The preliminary results indicate a notable influence of the pH and the elevation
over the island's microbiota. In particular, sites in higher altitudes found to
be inhabited by a more diverse number of microorganisms, a pattern commonly
seen in several faunistic groups, such as arthropods. These data along with the
open access data regarding arthropod fauna, flora and the vast data of remote
sensing and biodiversity, provide the basis to identify major drivers of
biodiversity, to evaluate hotspots and contribute to foreknowledge of threatened ecosystems.}

@@

The complilation of Crete data cube has multiple spatial data layers of global,
european, greek or cretan scale. 
The Copernicus CORINE Land Cover has 3 layers resolution that classify the land
use and cover in shapefile format \cite{CLC2023}. 
WorldClim 2.0 contains global climatic data for 12 variables, e.g annual mean
temperature and annual precipitation \cite{Fick2017}.
We also utilised teh Environmentally Sensitive Areas Index to desertification (ESAI) 
of Greece dataset \cite{KARAMESOUTI2018266}. Additionally we included the 
Global Aridity Index and Potential Evapotranspiration Database \cite{zomer2022version}.
Crete geology, bedrock formations were downloaded from IGME.
Handling and analysis of these data was done with the sf and stars R packages \cite{Pebesma2023}.
%World Database on Protected Areas (WDPA) \cite{Hanson2022}
%Average NDVI, 2000-2020
%Harmonized World Soil Database
\subsection{Network inferrence}\label{network_inferrence}

Network inferrence using FlashWeave 0.19.2 \cite{Tackmann2019}
Network analysis in igraph R package \cite{Csardi2006}
Julia Version 1.9.3 \cite{Julia-2017}

\subsection{Tools}\label{tools}
Function annotation - FAPROTAX \cite{Louca2016}
%PEMA for OTU inferrence \cite{Zafeiropoulos2020}
%ape R packege \cite{Paradis2004}
UMAP ordination \cite{mcinnes2018umap-software}
Numerical ecology analyses with vegan R package \cite{vegan}

Python 3.11.4
R version 4.3.2 \cite{rcoreteam}
Differential abundance analysis with ANCOM-BC R package \cite{Lin2020,Lin2023}
U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}
ggplot2 \cite{ggplot22016}
pheatmap \cite{Kolde2019}

Computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.

>>>>

\abstract{
    Microbes are known for their versatility, abundance
    and influence on soil ecosystem functioning.
    A synthesized knowledge base of microbial biodiversity, in terms of
    ecological and remote-sensing data remains a major challenge.
    Many worldwide studies have been published regarding soil
    microbiome ecosystems, though there are still many blind spots.
    Islands can be important case studies for this integration for more resolute and dense samplings.
    Here, we utilize the Island Sampling Day Crete 2016 microbial 16S rRNA gene
    amplicon data, integrated with soil and remote
    sensing data, to decipher the drivers of ecosystem function of the island.
    The Island Sampling Day Crete 2016 project has collected 144 topsoil samples
    from 72 sites, capturing a lot of this diversity, accompanied by FAIR
    (Findable, Accessible, Interoperable and Reproducible) data by design. 
    Cretan macroecology has been studied for centuries for its diverse  and endemic
    fauna and flora.
    In addition, Crete has been considered as a miniature continent with high contrasts in
    vegetation cover, elevation, climatic conditions. 
    We show that, higher altitudes in Crete found to
    be inhabited by a more diverse number of microorganisms, a pattern commonly
    seen in several faunistic groups, such as arthropods.
    The integration of the spatial data presented here enabled warning signals
    in pristine and grazing ecosystems.
    These results along with the
    climatic and desertification index influences on the soil microbiome of Crete,
    provide the basis to identify major drivers of biodiversity, to evaluate hotspots
    and contribute to foreknowledge of threatened ecosystems.
}

@@

The complilation of Crete data cube has multiple spatial data layers of global,
european, greek or cretan scale. 
The Copernicus CORINE Land Cover has 3 layers resolution that classify the land
use and cover in shapefile format \cite{CLC2023}. 
WorldClim 2.0 contains global climatic data for 12 variables, e.g annual mean
temperature and annual precipitation \cite{Fick2017}.
We also utilised teh Environmentally Sensitive Areas Index to desertification (ESAI) 
of Greece dataset \cite{KARAMESOUTI2018266}. Additionally we included the 
Global Aridity Index and Potential Evapotranspiration Database \cite{zomer2022version}.
Crete geological bedrock formations were downloaded from IGME.
Handling and analysis of these data was done with the sf and stars R packages \cite{Pebesma2023}.
%World Database on Protected Areas (WDPA) \cite{Hanson2022}
%Average NDVI, 2000-2020
%Harmonized World Soil Database
\subsection{Network inferrence}\label{network_inferrence}

The network inferrence was facilitated with FlashWeave 0.19.2 \cite{Tackmann2019}
in Julia language version 1.9.3 \cite{Julia-2017}. To use FlashWeave we reduced 
the abandunce table of the ASVs to keep the ones at the genus or species level.
In addition, we filtered these taxa that appeared in ### samples and had more than 
#### mean relative abundance. The subsequent network analyses were carried out
with the igraph R package \cite{Csardi2006}.

\subsection{Tools}\label{tools}
%PEMA for OTU inferrence \cite{Zafeiropoulos2020}
%U-CIE R package for coloring 3 dimentional data \cite{Koutrouli2022}

For taxa function annotation we used the manually curated FAPROTAX database and python script \cite{Louca2016}.
Whereas for the differential abundance analysis with ANCOM-BC2 R package \cite{Lin2023}.
Numerical ecology analyses,e.g diversity indices, NMDS, PERMANOVA we calculated
with the vegan R package \cite{vegan}.
For PCoA ordination we used ape R packege \cite{Paradis2004} and UMAP python library\cite{mcinnes2018umap-software}.
Visualisation was implemented with ggplot2 \cite{ggplot22016} and pheatmap \cite{Kolde2019}.
The environment we worked had Python 3.11.4 and R version 4.3.2 \cite{rcoreteam}. 
Finally, computations were performed on HPC infrastrure of HCMR \cite{Zafeiropoulos2021}.
